---
title: "mbQTL_Vignette"
date: "2023-02-05"
author: "Mercedeh Movassagh"
package: mbQTL
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{MbQTL_Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r style, echo=FALSE, results='asis'}
BiocStyle::markdown()
```


```{css, echo=FALSE}
.header-section-number {
  display: none;
}

body {
  counter-reset: counter-level-1;
}

h1:not(.title) {
  counter-increment: counter-level-1;
  counter-reset: counter-level-2;
}

h1:not(.title)::before{
  content: counter(counter-level-1, upper-alpha) ". ";
}

h2 {
  counter-incrementx: counter-level-2;
  counter-reset: counter-level-3;
}

h2::before {
  content: counter(counter-level-1, upper-alpha) "." counter(counter-level-2) " ";
}

h3 {
  counter-increment: counter-level-3;
}

h3::before {
  content: counter(counter-level-1, upper-alpha) "." counter(counter-level-2) "." counter(counter-level-3) " ";
}
```

```{r setup}
knitr::opts_chunk$set(
  dpi = 300,
  warning = FALSE,
  collapse = TRUE,
  error = FALSE,
  comment = "#>",
  echo = TRUE
)
library(mbQTL)
library(tidyr)
data("SnpFile")
data("microbeAbund")
data("CovFile")
data("metagenomeSeqObj")

doctype <- knitr::opts_knit$get("rmarkdown.pandoc.to")
```

```{r}
sessionInfo()
```

# Introduction

`mbQTL` is a statistical package for identifying significant taxa (microbial), 
snp (genotype) relationships across large sample sets. The package measures 
this relationship using three different approaches. A) linear regression analysis
which uses core `MatrixeQTL` model. B) rho estimation for correlation of taxa and SNP,
which is a novel method of estimating rho values after R value of taxa-taxa is 
estimated then adjusted for association of taxa-SNP. Using this method various
taxa-SNP associations can be estimated simultaneously, and linkage disequilibrium (LD)
snps and regions can be identified in terms of presence and their associations with
snps (see section for more details). C) Utilizing Logistic regression analysis, we
estimate the relationship of one snp and one taxa across the sample cohort but
simultaneously this could be estimated for all samples and proper statistics is
produced from these regressions. Appropriate plots are generated for all three
methods utilized in this vignette and are part of the package to simplify 
visualization.

# Data upload.


The datasets used for this vignette are all simulated and are not factual. The "Taxa_table_MVP_cor.txt"
and "SNP_table_MVP_cor.txt" are used in all three sections mentioned above and the "covariates.txt" 
file is used for section A)linear regression. The input file format for "SNP_table_MVP_cor.txt" 
should be colnames with SNP names/locations/rs_numbers and rownames needs to be sample number.
For the "Taxa_table_MVP_cor.txt" the colnames is the microbe name and the row name is the sample
number and finally for the "covariates.txt" file the colnames need to be samplenames and the 
rownames need to be the covariates. Please ensure all 3 files are matching in terms of sample
number and are appropriately formatted to avoid errors in your analysis.

## Import data
mbQTL accepts data in dataframe, `MRexperiment` objects or table formats. The files 
are accessible when the `mbQTL` library is loaded. 

```{r eval=TRUE, echo=TRUE}
# microbial count file
# microbeAbund
# SNP file
# SnpFile
# Covariance file
# CovFile
metagenomeSeqObj
```


If the user has an MRexperiment microbiome file (`metagenomeSeq` formatted), one can
use the following function to normalize their `metagenomeSeq` object or aggregate based
on various taxonomic levels before normalization. 
The output will be formatted in compatible data frame format for `mbQTL`. If one has
a `phyloseq` S4 object, they can use the `phyloseq_to_metagenomeSeq()` option of the
`phyloseq` package and start from there.


```{r eval=TRUE, echo=TRUE}
compatible_metagenome_seq <- metagenomeSeqToMbqtl(metagenomeSeqObj,
  norm = TRUE, log = TRUE,
  aggregate_taxa = "Genus"
)

# Check for the class of compatible_metagenome_seq

class(compatible_metagenome_seq)
```


# Part A. Lineare regression

Linear regression analysis to identify significant association between taxa and 
SNPs across large sample sized. P values FDRs, Pvalue histogram plots and QQ-plots
are provided for various taxa and SNP regression analysis performed. microbeAbund
is the taxa abundance file, SnpFile us file across samples (0,1,2  (reference,
heterozygous and alternate genotype). `MatrixeQTL` makes its estimations assuming
three different models of which we employ two, a) assumption of simple linear
regression for each gene and SNV pair The first approach is a very similar approach
to that of the expression quantitative trait loci association (eQTL). b) regression
model with covariates (Shabalin, 2012).


```{r eval=TRUE, echo=TRUE}

# perform linear regression analysis between taxa and snps across 
# samples and produce a data frame of results.(covariate file is
# optional but highly recommended to avoid results which might be
# prone to batch effects and obtaining optimal biological relevance
# for finding snp - taxa relationships.) The input files are 
# microbial abundance file (rownames taxa, colnames sample names) and
# SnpFile (rownames snps and colnames sample names) with optional "CovFile"
# (rownames covariates and colnames sample names) ( note all samples should
# be concordant in the three datasets).

LinearAnalysisTaxaSNP <- linearTaxaSnp(microbeAbund,
  SnpFile,
  Covariate = CovFile
)
```


```{r eval=TRUE, echo=TRUE, fig.width = 4, fig.height=4, out.width="90%", dpi=300, fig.align="center"}

# Create a histogram of P values across all snp - taxa
# linear regression estimations

histPvalueLm(LinearAnalysisTaxaSNP)


# Create a QQPlot of expected versus estimated P value for all all
# snp - taxa linear regression estimations

qqPlotLm(microbeAbund, SnpFile, Covariate = CovFile)
```


# Part B. Correlation analysis (rho estimation)

Estimate taxa-taxa correlation (microbeAbund (taxa abundance file)) and estimate R 
(Strength of the relationship between taxa `x` and `y`). In other words we estimate 
Goodness of fits, as a measure of a feature’s association with host genetics 
(Significant SNPs) association with taxa (matching 16S results). In details we
measure the strength of relationship between taxa `x` and taxa `y` (R correlation value),
next we measure goodness of fits R2 and ask to what extent the variance in one taxa
explains the variance in a SNP,  and finally we find clusters of taxa associated with
various SNPs across the dataset by evaluating and controlling for both these measures
together and finally estimating rho value.


```{r eval=TRUE, echo=TRUE}

# Estimate taxa SNP - taxa correlation R and estimate R2
# (To what extent variance in one taxa explains the
# variance in snp)

# First estimate R value for all SNP-Taxa relationships

correlationMicrobes <- coringTaxa(microbeAbund)

# Estimate the correlation value (R2) between a specific snp and all taxa.

one_rs_id <- allToAllProduct(SnpFile, microbeAbund, "chr1.171282963_T")

# Estimate the correlation value (R2) between all snps and all taxa.

for_all_rsids <- allToAllProduct(SnpFile, microbeAbund)

# Estimate rho value for all snps and all taxa present in the dataset.

taxa_SNP_Cor <- taxaSnpCor(for_all_rsids, correlationMicrobes)

# Estimate rho value for all snps and all taxa present in the
# dataset and pick the highest and lowest p values in the
# dataset to identify improtant SNP-Taxa relationships.

taxa_SNP_Cor_lim <- taxaSnpCor(for_all_rsids,
  correlationMicrobes,
  probs = c(0.0001, 0.9999)
)
```


```{r eval=TRUE, echo=TRUE, fig.width = 9, fig.height=5, out.width="90%", dpi=300, fig.align="center"}

## Draw heatmap of rho estimates "taxa_SNP_Cor_lim" is the
# output of taxaSnpCor().
# One can use other pheatmap() settings for extra annotation

mbQtlCorHeatmap(taxa_SNP_Cor_lim,
  fontsize_col = 5,
  fontsize_row = 7
)
```


# Part C. Logistic Regression analysis
SNPs and then Genus/species (categorical(presence/absence)) as expression. For this we need to binarize out taxa abundance file based on a cutoff to a zero or one or presence or absence format. We assume a presence and absence binary/categorical relationship between every taxa and SNP pair and use a logistic regression model to identify significance taxa-SNP relationships.

```{r eval=TRUE, echo=TRUE}

# perform Logistic regression analysis between taxa and snps 
# across samples microbeAbund is the microbe abundance file
# (a dateframe with rownames as taxa and colnames and sample
# names) and SnpFile is the snp file (0,1,2) values
# (rownames as snps and colnames as samples)

log_link_resA <- logRegSnpsTaxa(microbeAbund, SnpFile)

# Perform Logistic regression for specific microbe

log_link_resB <- logRegSnpsTaxa(microbeAbund, SnpFile,
  selectmicrobe = c("Haemophilus")
)
```



```{r eval=TRUE, echo=TRUE, fig.width = 5, fig.height=2, out.width="90%", dpi=300, fig.align="center"}

# Create a barplot with the specific rsID, and microbe of interest,
# including the genotype information for the reference versus 
# alternate versus hetrozygous alleles and and presence absence
# of microbe of interest. Note your reference, alternative and
# heterozygous genotype values need to match the genotype of
# your SNP of interest this information can be obtained
# from snpdb data base or GATK/plink files.

Logit_plot <- logitPlotSnpTaxa(microbeAbund, SnpFile,
  selectmicrobe = "Neisseria",
  rsID = "chr2.241072116_A", ref = "GG", alt = "AA", het = "AG"
)

Logit_plot
```
